---
title: "Reproducible Research: Peer Assessment 1"
author:  "Melissa Cuaycong"
output:  
  html_document:   
    keep_md: true
---


## Loading and preprocessing the data
```{r ReadData}
activity<-read.csv("./activity/activity.csv", na.strings="NA")
```

## What is mean total number of steps taken per day?
```{r ProcessData_TotalSteps}
library(plyr)

#Calculate the total steps per day
stepsPD <-ddply(activity,.(date),summarize,
                stepspd = sum(steps))

hist(stepsPD$stepspd,main="Total Steps Per Day",
     xlab="Total Steps",
     ylab="Frequency")

#Calculate the mean of the total number of steps taken per day.
meanSteps <-format(mean(stepsPD$stepspd,na.rm=TRUE),big.mark=",",scientific=FALSE)
medianSteps <-format(median(stepsPD$stepspd,na.rm=TRUE),big.mark=",",scientific=FALSE)
```

>The MEAN of the total number of steps taken per day is _**`r meanSteps`**_.
The MEDIAN of the total number of steps taken per day is _**`r medianSteps`**_.

## What is the average daily activity pattern?
```{r ProcessData_AverageDaily}
#Calculate the average number of steps per interval across all days.
stepsInt <-ddply(activity, .(interval), summarize,
                 meansteps = mean(steps,na.rm=TRUE))

plot(stepsInt$interval,stepsInt$meansteps,type="l",
     xlab="5-minute Interval",
     ylab="Average Steps")
               
#Get the max of the average steps per interval.
maxSteps<-format(max(stepsInt$meansteps),big.mark=",",digits=1) 

#Determine which 5-minute interval this maximum occurs.
maxInt<-stepsInt[which.max(stepsInt$meansteps),"interval"]
maxInt<-paste0(trunc(maxInt/100),":",((maxInt/100)-trunc(maxInt/100))*100)

```

>On average, across all the days, the maximum steps of _**`r maxSteps`**_ occurs in
the 5-minute interval, at  _**`r maxInt`**_.

## Imputing missing values


```{r FindMissing}
#Calculate and report the total number of missing values in the dataset (i.e. the total
#number of rows with NAs) 

completerows<-complete.cases(activity)
numNArows <-format(table(completerows)["TRUE"],big.mark=",")
```

>There are _**`r numNArows`**_ rows with missing values.

>Create an adjusted activity dataset with the NA values for number of steps imputed.

>**Methodology:**  Use the average steps for the specific interval (calculated from available data across all days), to replace the NA values for the specific interval.
   
```{r ImputeMissing}
#Identify and split missing rows from dataset
narows<-activity[!complete.cases(activity),c("date","interval","steps")]
fullrows<-activity[complete.cases(activity),c("date","interval","steps")]

#Substitute NA values for specific interval with ff methodology 
#- Average steps for the specific interval across all days
newnarows<-merge(x=narows, y=stepsInt, by.x="interval")
newnarows<-newnarows[,c("date","interval","meansteps")]
names(newnarows)[names(newnarows)=="meansteps"] <- "steps"

#Create duplicate dataset with missing values imputed
activityAdj<-rbind(fullrows,newnarows)

```

Next, evaluate the adjusted dataset with the imputed NA values.
Compare findings with the unadjusted dataset.

```{r EvaluateNAAdjDataset}
#Calculate total steps by day for adjusted dataset.
stepsPDA <-ddply(activityAdj,.(date),summarize,
                stepspd = sum(steps))

#Create a histogram
hist(stepsPDA$stepspd, main="Total Steps Per Day (NAs Imputed)",
     xlab="Total Steps",
     ylab="Frequency")

#Calcualte the mean and median steps for total steps per day.
meanStepsA <-format(mean(stepsPDA$stepspd,na.rm=TRUE), big.mark=",",scientific=FALSE)
medianStepsA <-format(median(stepsPDA$stepspd,na.rm=TRUE),big.mark=",",scientific=FALSE)
```

>The MEAN of the total number of steps taken per day is _**`r meanStepsA`**_.
The MEDIAN of the total number of steps taken per day is _**`r medianStepsA`**_.
The MEAN and MEDIAN are the same for the adjusted dataset.

###Compare mean/median data between the 2 datasets.

>
Measure       | Activity (NAs)| Activity (NAs imputed)
------------- | ------------- | ----------------------
MEAN          | `r meanSteps` | `r meanStepsA` 
MEDIAN        | `r medianSteps`|`r medianStepsA`


## Are there differences in activity patterns between weekdays and weekends?

Create a factor variable in the adjusted dataset to indicate whether the measurement was taken
on a weekday or a weekend.  Evaluate results.

```{r EvaluateWeekdays}

#Create a flag (factor variable) in the adjusted dataset to indicate weekday or
#weekend #based on the date #measurement was taken
tempday <- weekdays(as.Date(activityAdj$date),abbreviate=TRUE)
activityAdj$dayFlag<-as.factor(ifelse(tempday %in% c("Sat","Sun"),"Weekend","Weekday"))

#Calculate the average steps per interval, grouped by the weekday indicator.
stepsWD <-ddply(activityAdj, .(dayFlag,interval), summarize,
                        meansteps = mean(steps,na.rm=TRUE))

#Plot the results in separate panels by the weekday grouping.

library(ggplot2)
p <- ggplot(stepsWD, aes(interval,meansteps)) 
p <- p +
        geom_line() +
        facet_grid(dayFlag~.) +
        xlab("5-min Interval") +
        ylab("Average Steps") +
        ggtitle("Average Steps Per 5-Min Interval ")
print(p)
```

> While the maximum of the average steps per interval is higher on a weekday,
Weekend activity is, on average more active throughout the day than Weekday activity. 
There seems to be a flurry of activity prior to 10 AM on weekdays, starting the day
earlier (around 5 AM) than on weekends. However, the activity slows down after 10 AM.

>On the other hand, peak activity for weekends also happens prior to 10 AM.  But,
activity continues to throughout the day at a higher pace than on weekdays.
